---
title: "612 project 1"
author: "Albina Gallyavova"
date: "2/17/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r libraries, echo=FALSE}
library(knitr)
library(recommenderlab)
library(tidyverse)
library(DT)
data(MovieLense)
```

## Intro  
The recommender system will suggest a movie to a user that he/she previously has not seen. We will use MovieLense dataset available in `recommenderlab` package.
  
## Data  
Per Gorakala and Usuelli,   
- movies that have been viewed only a few times may be biased due to the lack of data  
- similarly, users who rated only a few movies, might also be biased.  
As such, we need to select only relevant movies.  

```{r}
# code source: Building a recommendation system with R textbook
# users who rated at least 10 movies and movies that have been watched at least 10 times
ratings <- MovieLense[rowCounts(MovieLense) > 10, colCounts(MovieLense) > 10]

# take only 1 percent of such users/movies
min_movies <- quantile(rowCounts(ratings),0.99) # minimum number of movies per user
min_users <- quantile(colCounts(ratings),0.99) # minimum number of users per movie
r <- ratings[rowCounts(ratings) > min_movies,colCounts(ratings) > min_users] 

# heatmap of top users and movies
image(r)
```
  
## Train and test  
```{r}
set.seed(1)
r_df <- as(r,'data.frame')
train <- sort(sample(nrow(r_df),nrow(r_df)*.70))
train_df = r_df[train,]
test_df = r_df[-train,]

# user-item matrix - wide format
train_df <- reshape(train_df, idvar = "user", timevar = "item", direction = "wide")
test_df <- reshape(test_df, idvar = "user", timevar = "item", direction = "wide")

# set row index to user id
row.names(train_df) <- train_df$user
row.names(test_df) <- test_df$user

#remove user column after setting index
train_df <- within(train_df, rm(user))
test_df <- within(test_df, rm(user))

# test data was missing one column, in order to make it easy to compare later to baseline, 
# we align dimensions and column order in test data with train
test_df[,setdiff(colnames(train_df), colnames(test_df))] <- NA
test_df <- test_df[names(train_df)]

datatable(train_df, options = list(scrollX = "300px"))
datatable(test_df, options = list(scrollX = "300px"))
```
  
## Raw average for train  
```{r}
raw_avg <- mean(as.matrix(train_df), na.rm=TRUE)
```
Raw average is `r round(raw_avg,2)`.  

## RMSE and raw average for train and test  
```{r}
raw_avg_rmse_train <- sqrt(mean(((as.matrix(train_df)-raw_avg)^2), na.rm=TRUE))
raw_avg_rmse_test <- sqrt(mean(((as.matrix(test_df)-raw_avg)^2), na.rm=TRUE))
```
Raw average RMSE for train is `r round(raw_avg_rmse_train,2)`.  
Raw average RMSE for test is `r round(raw_avg_rmse_test,2)`.  
  
## Bias for each user and item in train  
```{r}
u_bias <- rowMeans(train_df,na.rm=TRUE) - raw_avg
i_bias <- colMeans(train_df,na.rm=TRUE) - raw_avg
u_bias_df <- cbind(train_df,u_bias) 
i_bias_df <- rbind(u_bias_df,i_bias)

datatable(round(u_bias_df,2) %>% select(u_bias),options = list(scrollX = "300px"))
datatable(t(tail(round(i_bias_df,2),1)), options = list(scrollX = "300px"))
```
  
## Baseline predictor from the raw average, and the appropriate user and item biases  
```{r}
baseline_df <- data.frame(matrix(NA, nrow = nrow(train_df), ncol = ncol(train_df)))
baseline_df <- raw_avg + replicate(ncol(train_df),u_bias) + t(replicate(nrow(train_df),i_bias))
colnames(baseline_df) <- colnames(train_df)

# clip ratings at 5 for values greater than 5 and 1 for values below 1
baseline_df[baseline_df>5] = 5
baseline_df[baseline_df<1] = 1
datatable(round(baseline_df,2),options = list(scrollX = "300px"))
```
  
## RMSE for the baseline predictors for training and test data  
```{r}
baseline_pred_rmse_train <- sqrt(mean(as.matrix((baseline_df-train_df)^2), na.rm=TRUE))
baseline_pred_rmse_test <- sqrt(mean(as.matrix((baseline_df-test_df)^2), na.rm=TRUE))
```
  
## Summary  
```{r}
perc_impr <- (1-baseline_pred_rmse_test/raw_avg_rmse_test)*100
```

Comparing RMSE between raw average and baseline predictor we see that baseline predictor improved RMSE from raw average by `r round(perc_impr,2)` percent so we'll use baseline prediction for recommendations.
Original data contained NAs for some users, and assuming NAs resulted from users not seeing a movie, we can recommend such movie to the user if the rating is above certain threshold. 
Setting threshold at >=4, we have the following user-item combinations for recommendation.

```{r}
bl <- rownames_to_column(as.data.frame(baseline_df),var='user')
bl <- gather(bl,'item','rating',-user) %>% 
  mutate(item = str_replace(item,'rating.',''))

rec <- merge(r_df,bl,by=c('user','item'),all = TRUE)
rec[(is.na(rec$rating.x)) & (rec$rating.y>=4),]
```

